Geometric models reveal the hidden structure of conceptual knowledge

### 2023-01-23 paper/main.tex
working on results section

How do we acquire conceptual knowledge? Memorizing course lectures or textbook
chapters by rote can lead to the superficial \textit{appearance} of
understanding the underlying content, but achieving true conceptual
understanding seems to require something deeper and richer. Does conceptual
understanding entail connecting newly acquired information to the scaffolding
of one's existing knowledge or experience~\citep{BlayEtal06,CaraMaho03,
ConsEtal16, DeacEtal04, SimoEtal04}? Or weaving a lecture's atomic
elements (e.g., its component words) into a structured network that describes
how those individual elements are related? Conceptual understanding could also
involve building a mental model that transcends the meanings of those
individual atomic elements by reflecting the deeper meaning underlying the
gestalt whole~\citep{Kint70, Macl05, ScotEtal07}.

@@

What form might the representation of the sum total of a person's knowledge
take? First, we might require a means of systematically describing or
representing the nearly infinite set of possible things a person could know.
Second, we might want to account for potential associations between different
concepts. For example, the concepts of ``fish'' and ``water'' might be
associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and should also
unlock any now-satisfied dependencies of that newly learned concept so that
they are ``tagged'' as available for future learning.

Here we develop a framework for modelling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes the extent to which each concept could be learned. Each location
on these maps represents a single concept, and the geometries are defined such
that related concepts are located nearby in space. We use this framework to
analyzing and interpreting behavioral data collected from an experiment that
has participants watch and answer conceptual questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can start
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning.

@@

>>>>

Suppose that a teacher had access to a complete ``map'' of everything a student
knew. Defining what such a map might even look like, let alone how it might be
constructed or filled in, is itself a non-trivial problem. But if a teacher
\textit{were} to gain access to such a map, how might that change their ability
to teach the student? Perhaps they might start by checking how well the student
knew the to-be-learned information already, or how much they knew about related
concepts. For some students, they could potentially optimize their teaching
efforts to maximize efficiency by focusing primarily on not-yet-known content.
For other students (or other content areas), it might be more effective to
optimize for direct connections between already-known content and any new
material. Observing how the student's knowledge was changing over time, in
response to their training, could also help to guide the teacher.

Designing and building procedures and tools for mapping out knowledge touches
on deep questions about what it means to learn. For example, how do we acquire
conceptual knowledge? Memorizing course lectures or textbook chapters by rote
can lead to the superficial \textit{appearance} of understanding the underlying
content, but achieving true conceptual understanding seems to require something
deeper and richer. Does conceptual understanding entail connecting newly
acquired information to the scaffolding of one's existing knowledge or
experience~\citep{BlayEtal06,CaraMaho03, ConsEtal16, DeacEtal04, SimoEtal04}?
Or weaving a lecture's atomic elements (e.g., its component words) into a
structured network that describes how those individual elements are related?
Conceptual understanding could also involve building a mental model that
transcends the meanings of those individual atomic elements by reflecting the
deeper meaning underlying the gestalt whole~\citep{Kint70, Macl05, ScotEtal07}.

@@

Given these insights, what form might the representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and should also
unlock any now-satisfied dependencies of that newly learned concept so that
they are ``tagged'' as available for future learning.

Here we develop a framework for modelling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes the extent to which each concept could be learned. Each location
on these maps represents a single concept, and the geometries are defined such
that related concepts are located nearby in space. We use this framework to
analyze and interpret behavioral data collected from an experiment that has
participants watch and answer conceptual questions about a series of recorded
course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can start
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a ``proof of concept''
demonstration of how our models might be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing someone's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track our participants' knowledge and how it changed over time in response
to training.

We asked our participants to answer questions from several multiple choice
quizzes and watch two lecture videos from the \textit{Khan Academy} platform
(Fig.~\ref{fig:experiment}). One lecture video, entitled \textit{Four
Fundamental Forces}, was about the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second lecture video,
entitled \textit{Birth of Stars}, provides an overview of our current
understanding of how stars form. We selected both lessons to be (a) accessible
to a broad audience, e.g., by minimizing prerequisite knowledge, (b) largely
independent of each other, e.g., so that the two videos focused on different
material and did not depend on each other, and (c) related to each other, e.g.,
so that both videos contained at least \textit{some} similar or overlapping
content. The two videos we selected are introductory, about different primary
concepts, but also touch on ``physics'' and ``astronomy'' themes. We also wrote
a set of multiple choice quiz questions that would enable us to test
participants' knoweldge about each individual video and about related content
not specifically presented in either video (Tab.~\questions). Participants
answered questions randomly drawn from each content area (lecture 1, lecture 2,
and general physics knowledge) across each of three quizzes. Quiz 1 was
intended to assessed participants' knowledge before training; quiz 2 assessed
knowledge after watching the Four Fundamental Forces video (i.e., lecture 1);
and quiz 3 assessed knowledge after watching the Birth of Stars video (i.e.,
lecture 2).

We trained a text embedding model using sliding windows of text from the two
videos' transcripts (see \textit{Constructing text embeddings of multiple
videos and questions}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single topic
vector-- i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/lecture-question-similarity}

    \caption{\textbf{Which parts of each lecture are captured by each
    question?} Each panel displays timeseries plots showing how each question's
    topic vector correlates with each video timepoint's topic vector. The left
    panel displays these correlations for the \textit{Four Fundamental Forces}
    lecture and associated questions, and the right panels displays these
    correlations for the \textit{Birth of Stars} lecture and associated
    questions. The colors denote question identities. The diamonds in each
    panel denote the moment of peak correlation between the indicated
    questions, in the indicated lectures. The associated questions' text, and
    snippets of the lectures' transcripts in the best-matching sliding windows,
    are displayed at the bottom of the figure.}

    \label{fig:question-correlations}
\end{figure}

Although a single lecture may be organized a single broad theme at a coarse
scale, at a finer scale each moment of a lecture typically covers a narrower
range of content. We wondered whether a text embedding model trained on the
lectures' transcripts might capture some of this finer scale content. For
example, if a particular question asks about the content from one small part of
a lecture, we wondered whether our text embedding model could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined range of timepoints
from their corresponding lectures, and the correlations fell off sharply
outside of that range. We also examined the best-matching intervals for each
question qualitatively by comparing the text of the question to the text of the
most-correlated parts of the lectures. Despite that the questions were excluded
from the text embedding model's training set, in general we found a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Fig.~\ref{fig:question-correlations}.

The ability to quantify how much each question is ``asking about'' each moment
of the lectures could enable high-resolution insights into participants'
knowledge. Traditional approaches to estimating how much a student ``knows''
about the content of a given lecture entail computing the proportion of
correctly answered questions. But if two students receive identical scores on
an exam, might our modeling framework help us to gain more naunced insights
into the \textit{specific} content that each student has mastered (or failed to
master)? For example, a student who misses three questions that were all about
the same concept (e.g., concept $A$) will have gotten the same
\textit{proportion} of questions correct as another student who missed three
questions about three \textit{different} concepts (e.g., $A$, $B$, and $C$).
But if we wanted to fill in the ``gaps'' in the two students' understandings,
we might do well to focus on concept $A$ for the first student, but add in
concepts $B$ and $C$ for the second student.

\textbf{JRM STOPPED HERE...}

As a sanity check, we anticipated that our knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each video.
